---
title: "Phyloseq demo"
author: "R Vannette"
date: "June 29, 2015"
output: html_document
---

## Load packages and data

```{r}
#getwd()
#setwd() # change this to the location of your data


# load packages
#install.packages("phyloseq")
#install.packages("ape")
library(phyloseq)
library(ape)

# get files
# sample data
samp_dat <- read.csv("data/Centralia_Full_Map.txt", 
                       sep = "\t", header = T )

rownames(samp_dat) <- samp_dat$X.SampleID #row names must match OTU table headers

SAMP <- sample_data(samp_dat)

# biom file
biom <- import_biom("data/otu_table_mc2_json.biom", header=T)
OTU <- otu_table(biom, taxa_are_rows = TRUE)

```

Note: the .biom table should be in json format. See
<http://biom-format.org/> for more details. 

If you're working with mothur output files, phyloseq has the `import_mothur` command that allows the user to import the shared/otu table, taxonomy, and phylogenetic tree files into phyloseq in one command.


```{r}
# taxonomy file
taxonomy<-read.table("data/rep_set_tax_assignments.txt", sep="\t", header=F)

###
#code to split up the taxonomy string #
##

tax<-as.character(taxonomy$V2)
head(tax)
#root<-NA
Kingdom<-NA
Phylum<-NA
Class<-NA
Order<-NA
Family<-NA
Genus<-NA
Species<-NA

linaean<-data.frame(Kingdom,Phylum,Class,Order,Family,Genus,Species)

for (i in 1:dim(taxonomy)[1]) {
  
  split.tax<-strsplit(tax[i],split=c(";"))
  linaean[i,1]<-substr(split.tax[[1]][1], 4,30)
  linaean[i,2]<-substr(split.tax[[1]][2], 4,30)
  linaean[i,3]<-substr(split.tax[[1]][3], 4,30)
  linaean[i,4]<-substr(split.tax[[1]][4], 4,30)
  linaean[i,5]<-substr(split.tax[[1]][5], 4,30)
  linaean[i,6]<-substr(split.tax[[1]][6], 4,30)
  linaean[i,7]<-substr(split.tax[[1]][7], 4,30)
}
##

# remove underscore before taxon name
rownames(linaean) <- taxonomy$V1
for(i in 2:7){
linaean[,i] <- gsub("_","",linaean[,i])
}

TAX <- tax_table(as.matrix(linaean))

##if you want to import representative sequences of your OTUs which is basically your rep_set.fna file after you have picked OTUs in QIIME

#import ref seq
refseqs<-readDNAStringSet("rep_set.fna")
refseqs
head(names(refseqs))

#parse names of each sequence such that they match the format of OTU names in other files
#use substitution function, for these, see the biom table in csv format and find
#out the the length of the headers before space, in my example, the shortest id is 5 and longest is 28 before the header starts
names(refseqs) = gsub("(^[[:print:]]{5,28})([[:space:]]{1,})([[:print:]]{0,}$)","\\1", names(refseqs))
head(names(refseqs))
tail(names(refseqs))


# tree file
tre<-read.tree("data/rep_set.tre")
# tree tip labels must match OTU names above. 
```

## Combine the data into a phyloseq-class object

Note that phyloseq works with a **phyloseq-class object**, which is an object that has 4 types of data inside of it:  
1. OTU Table:  Known as `otu_table()`  
2. Taxonomy File:  Known as `tax_table()`  
3. Metadata File with sample information: Known as `sample_data()`  
4. Phylogenetic Tree File:  Known as `phy_tree()`

```{r}
# Put the 4 data types together into a phyloseq-class object
myphy<- phyloseq(OTU, TAX, SAMP, tre)

```

##Data filtering and subsampling

Filter out OTUs that do not match Kingdom Bacteria

```{r}
dim(otu_table(myphy)) #check dimensions of the otu table
myphy_f <- subset_taxa(myphy, Kingdom=="Bacteria") #include all OTUS identified as Bacteria
dim(otu_table(myphy_f)) #check how many were excluded

myphy_f <- subset_taxa(myphy_f, Class!="Chloroplast") #exclude all OTUS identified as Chloroplast
dim(otu_table(myphy_f)) #check how many were excluded



```

Normalize/rarefy OTU table (once)
```{r}
myphy_f_rare <- rarefy_even_depth(myphy_f, rngseed=1) #record seed and # of OTUs removed

```

##Estimate and plot within-sample diversity

```{r}
library(ggplot2)

plot_richness(myphy_f_rare, x="Classification", 
              measures=c("Observed", "Shannon"))+
  geom_boxplot(aes(color=Classification))+
  theme_bw()

```

##Comparative diversity and ordination

```{r}
library(vegan)

GPdist = phyloseq::distance(myphy_f_rare, "bray") # create bray-curtis distance matrix
GP.ord <- ordinate(myphy_f_rare, "NMDS", "bray") # NMDS ordination 

plot_ordination(myphy_f_rare, ordination = GP.ord, type = "samples", 
                color = "Classification",   title = "Bacterial NMDS (Bray-Curtis dissim.)")+theme_bw()

adonis(GPdist~Classification,  as(sample_data(myphy_f_rare), "data.frame")) #Are there significant changes? 

```




*Combine all taxa assigned to the same Class for these analysis, 
otherwise plotting the tree and barplot can take a long time. *

Here, we combine all OTUs with the same taxonomic classification at the level of Class


```{r}
dim(otu_table(myphy_f_rare)) #check dimensions of the otu table

myphy_Class <- tax_glom(myphy_f_rare, taxrank="Class") #agglomerate taxa based on class level

dim(otu_table(myphy_Class)) 

```


## Composition barplots

```{r}

plot_bar(myphy_Class, fill="Phylum") 

```

Now merge replicates and re-make plots

```{r}
myphy_Class_merged <- merge_samples(myphy_Class, "Sample")

plot_bar(myphy_Class_merged, fill = "Phylum") 
```

## Phylogenetic tree

```{r}
quartz()
plot_tree(myphy_Class_merged, 
          color = "Phylum", 
          label.tips = "Class", 
          base.spacing = 0.01,
          text.size = 3,
          size = "Abundance")+
  guides(color=guide_legend(ncol=2))

```


## Subsetting made easy

```{r}

myphy_thermo <- subset_taxa(myphy_f_rare, Class=="Thermoleophilia")
quartz()
plot_tree(myphy_thermo, 
          color = "Classification",  
          #label.tips = "Family",
          base.spacing = 0.01,
          text.size = 2,
          size = "Abundance")+
  guides(color=guide_legend(ncol=2))

```


## Other resources

Phyloseq main page with links to tutorials: 
<https://joey711.github.io/phyloseq/>

RDP tutorials with phyloseq: 
 https://rdp.cme.msu.edu/tutorials/stats/using_rdp_output_with_phyloseq.html



